---
title: "Hands-on Exercise 4 -  Funnel Plots for Fair Comparisons"
author: "Alicia Loh"
date: "April 30, 2024"
date-modified: last-modified
execute: 
  eval: true
  echo: true
  warning: false
  freeze: false
---

Funnel plot is a specially designed data visualisation for conducting unbiased comparison between outlets, stores or business entities. 

Learning Objectives:

-   plotting funnel plots by using **funnelPlotR** package,

-   plotting static funnel plot by using ggplot2 package, and

-   plotting interactive funnel plot by using both **plotly R** and **ggplot2** packages.

# Getting Started

## Installing and loading the required libraries

The following R packages will be used:

-   **readr** for importing csv into R.

-   **FunnelPlotR** for creating funnel plot.

-   **ggplot2** for creating funnel plot manually.

-   **knitr** for building static html table.

-   **plotly** for creating interactive funnel plot.

Code chunk below will be used to check if these packages have been installed and also will load them into the working R environment.

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

## Importing the Data

The COVID-19_DKI_Jakarta will be used. The data was downloaded from [Open Data Covid-19 Provinsi DKI Jakarta portal](https://riwayat-file-covid-19-dki-jakarta-jakartagis.hub.arcgis.com/).

For this hands-on exercise, compares the cumulative COVID-19 cases and death by sub-district (i.e. kelurahan) as at 31st July 2021, DKI Jakarta.

The code chunk below imports the data into R and save it into a tibble data frame object called *covid19*.

```{r}
covid19 <- read_csv("data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

# **FunnelPlotR methods**

A point estimate is a single number, e.g., mean. Uncertainty, is expressed as standard error, confidence interval, or credible interval.

The code chunk below will be used to derive the necessary summary statistics.

-   `group_by()` of **dplyr** package is used to group the observation by RACE,

-   `summarise()` is used to compute the count of observations, mean, standard deviation

-   `mutate()` is used to derive standard error of Maths by RACE, and

-   the output is save as a tibble data table called *my_sum*.

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
    ) %>%
  mutate(se=sd/sqrt(n-1))
```

The code chunk below will be used to display *my_sum* tibble data frame in an html table format.

::: panel-tabset
## Table

```{r, echo=FALSE}

knitr::kable(head(my_sum), format = 'html')
```

## Code

```{r, eval=FALSE}

knitr::kable(head(my_sum), format = 'html')
```
:::

## **Plotting standard error bars of point estimates**

The code chunk belows plots the standard error bars of mean maths score by race.

Note:

-   The error bars are computed by using the formula mean+/-se.

-   For `geom_point()`, it is important to indicate *stat=“identity”*.

::: panel-tabset
## Plot

```{r, echo=FALSE}

ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by rac")
```

## Code

```{r, eval=FALSE}

ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by rac")
```
:::

## **Plotting confidence interval of point estimates**

Instead of plotting the standard error bar of point estimates, the confidence intervals of mean maths score by race can also be plotted.

Note:

-   The confidence intervals are computed by using the formula mean+/-1.96\*se.

-   The error bars is sorted by using the average maths scores.

-   `labs()` argument of ggplot2 is used to change the x-axis label.

::: panel-tabset
## Plot

```{r, echo=FALSE}

ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean), 
        ymin=mean-1.96*se, 
        ymax=mean+1.96*se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")             
```

## Code

```{r, eval=FALSE}
                                  
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean), 
        ymin=mean-1.96*se, 
        ymax=mean+1.96*se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    size=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```
:::

## **Visualizing the uncertainty of point estimates with interactive error bars**

The code chunk below plots interactive error bars for the 99% confidence interval of mean maths score by race.

::: panel-tabset
## Plot

```{r, echo=FALSE}

shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```

## Code

```{r, eval=FALSE}

shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```
:::

## **Visualising Uncertainty: ggdist package**

-   [**ggdist**](https://mjskay.github.io/ggdist/) is an R package that provides a flexible set of ggplot2 geoms and stats designed especially for visualising distributions and uncertainty.

-   It is designed for both frequentist and Bayesian uncertainty visualization, taking the view that uncertainty visualization can be unified through the perspective of distribution visualization:

    -   for frequentist models, one visualises confidence distributions or bootstrap distributions (see vignette(“freq-uncertainty-vis”));

    -   for Bayesian models, one visualises probability distributions (see the tidybayes package, which builds on top of ggdist).

## **Visualizing the uncertainty of point estimates: ggdist methods**

In the code chunk below, [`stat_pointinterval()`](https://mjskay.github.io/ggdist/reference/stat_pointinterval.html) of **ggdist** is used to build a visual for displaying distribution of maths scores by race.

::: panel-tabset
## Plot

```{r, echo=FALSE}

exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval() +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")

```

## Code

```{r, eval=FALSE}

exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval() +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```
:::

In the code chunk below the following arguments are used:

-   .width = 0.95

-   .point = median

-   .interval = qi

::: panel-tabset
## Plot

```{r, echo=FALSE}

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(.width = 0.95,
  .point = median,
  .interval = qi) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot")
```

## Code

```{r, eval=FALSE}

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(.width = 0.95,
  .point = median,
  .interval = qi) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot")
```
:::

The plot below shows 95% and 99% confidence intervals

-   **`stat_pointinterval`** is used twice, once for each confidence interval.

-   The **`.width`** argument specifies the width of the intervals.

-   The **`.point`** argument specifies that we want to plot the median.

-   The **`.interval`** argument is set to "quantile" to indicate quantile-based intervals.

-   **`scale_colour_manual`** is used to set custom colors for the confidence intervals and provide custom labels.

-   Other aesthetic adjustments are made to improve the appearance of the plot, such as adjusting the size and position of the intervals.

::: panel-tabset
## Plot

```{r, echo=FALSE}

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(
    .width = 0.95,
    .point = "median",
    .interval = "quantile",
    aes(colour = "95% CI")) +
  stat_pointinterval(
    .width = 0.99,
    .point = "median",
    .interval = "quantile",
    aes(colour = "99% CI")) +
  scale_colour_manual(
    values = c("95% CI" = "blue", "99% CI" = "red"),
    labels = c("95% CI", "99% CI")) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot") +
  theme_minimal()

```

## Code

```{r, eval=FALSE}

exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(
    .width = 0.95,
    .point = "median",
    .interval = "quantile",
    aes(colour = "95% CI")) +
  stat_pointinterval(
    .width = 0.99,
    .point = "median",
    .interval = "quantile",
    aes(colour = "99% CI")) +
  scale_colour_manual(
    values = c("95% CI" = "blue", "99% CI" = "red"),
    labels = c("95% CI", "99% CI")) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot") +
  theme_minimal()
```
:::

## **Visualizing the uncertainty of point estimates: ggdist methods**

::: panel-tabset
## Plot

```{r, echo=FALSE}

exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval(
    show.legend = FALSE) +   
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

## Code

```{r, eval=FALSE}

exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval(
    show.legend = FALSE) +   
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```
:::

## **Visualising the uncertainty of point estimates: ggdist methods**

In the code chunk below, [`stat_gradientinterval()`](https://mjskay.github.io/ggdist/reference/stat_gradientinterval.html) of **ggdist** is used to build a visual for displaying distribution of maths scores by race.

::: panel-tabset
## Plot

```{r, echo=FALSE}

exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_gradientinterval(   
    fill = "skyblue",      
    show.legend = TRUE     
  ) +                        
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

## Code

```{r, eval=FALSE}

exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_gradientinterval(   
    fill = "skyblue",      
    show.legend = TRUE     
  ) +                        
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```
:::

## **Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)**

::: panel-tabset
## Plot

```{r, echo=FALSE}

ggplot(data = exam, 
       (aes(x = factor(RACE), y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, width = 0.05), 
    size = 0.4, color = "#0072B2", alpha = 1/2) +
  geom_hpline(data = sampler(25, group = RACE), height = 0.6, color = "#D55E00") +
  theme_bw() + 
  # `.draw` is a generated column indicating the sample draw
  transition_states(.draw, 1, 3)
```

## Code

```{r, eval=FALSE}

ggplot(data = exam, 
       (aes(x = factor(RACE), y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, width = 0.05), 
    size = 0.4, color = "#0072B2", alpha = 1/2) +
  geom_hpline(data = sampler(25, group = RACE), height = 0.6, color = "#D55E00") +
  theme_bw() + 
  # `.draw` is a generated column indicating the sample draw
  transition_states(.draw, 1, 3)
```
:::

## **Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)**

::: panel-tabset
## Plot

```{r, echo=FALSE}

ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)
```

## Code

```{r, eval=FALSE}

ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)
```
:::
