---
title: "Hands-on Exercise 2 - Beyond ggplot2 Fundamentals"
author: Alicia Loh
date-modified: last-modified
---

Learning Objectives:

-   Control the placement of annotation on a graph by using functions provided in ggrepel package,

-   Create professional publication quality figure by using functions provided in ggthemes and hrbrthemes packages,

-   Plot composite figure by combining ggplot2 graphs by using patchwork package.

# Getting Started

## Installing and loading the required libraries

Besides tidyverse, the following R packages will be used.

-   ggrepel: an R package provides geoms for ggplot2 to repel overlapping text labels.

-   ggthemes: an R package provides some extra themes, geoms, and scales for ‘ggplot2’.

-   hrbrthemes: an R package provides typography-centric themes and theme components for ggplot2.

-   patchwork: an R package for preparing composite figure created using ggplot2.

Code chunk below will be used to check if these packages have been installed and also will load them onto your working R environment.

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse)
```

## Importing the Data

-   The code chunk below imports *exam_data.csv* into R environment by using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package.

-   **readr** is one of the tidyverse package.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

**exam_data contains:**

-   Year end examination grades of a cohort of primary 3 students from a local school.

-   There are a total of seven attributes. Four of them are categorical data type and the other three are in continuous data type.

    -   The categorical attributes are: ID, CLASS, GENDER and RACE.

    -   The continuous attributes are: MATHS, ENGLISH and SCIENCE.

# **Beyond ggplot2 Annotation: ggrepel**

One of the challenge in plotting statistical graph is annotation, especially with large number of data points.

::: panel-tabset
## Plot

```{r}

#| echo: false

ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Code

```{r}
#| eval: false

ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

## **Working with ggrepel**

[**ggrepel**](https://ggrepel.slowkow.com/) is an extension of **ggplot2** package which provides `geoms` for **ggplot2** to repel overlapping text seen below. Simply replace `geom_text()` by [`geom_text_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel.html) and `geom_label()` by [`geom_label_repel`](https://ggrepel.slowkow.com/reference/geom_text_repel.html).

::: panel-tabset
## Plot

```{r}

#| echo: false

ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Code

```{r}
#| eval: false

ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

::: callout-note
Benefits of ggplot2 over in-built R graphics

-   Automated Legends: Unlike built-in plotting which requires manual legend creation, ggplot2 automatically generates legends based on the data's aesthetics. This streamlines the process, especially for beginners who may find managing legends challenging.

-   Simplified Faceting: ggplot2 simplifies the creation of faceted plots (multiple subplots) compared to base R plotting. The latter can be quite complex, requiring setup with par() and potentially confusing for loops, which might be overwhelming for beginners.

-   Tidy Data Integration: ggplot2 encourages the use of 'tidy data' principles, aligning with how other R tools and packages handle data. This practice helps beginners learn a consistent and widely-adopted approach to data analysis in R.
:::

# Grammar of Graphics

Grammar of Graphics is a general scheme for data visualization which breaks up graphs into semantic components e.g., scales and layers. It was introduced by Leland Wilkinson (1999) **Grammar of Graphics**, Springer. In summary, Grammar of Graphics defines the rules of structuring mathematical and aesthetic elements into a meaningful graph.

There are two principles in Grammar of Graphics, they are:

-   Graphics = distinct layers of grammatical elements

-   Meaningful plots through aesthetic mapping

A good grammar of graphics will allow us to gain insight into the composition of complicated graphics, and reveal unexpected connections between seemingly different graphics (Cox 1978). It also provides a strong foundation for understanding a diverse range of graphics. Furthermore, it may serve as a guide on what a well-formed correct graphic looks like.

## A Layered Grammar of Graphics

ggplot2 is an implementation of Leland Wilkinson’s Grammar of Graphics.

There are seven grammars of ggplot2:

-   **Data**: The dataset being plotted.

-   **Aesthetics** take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.

-   **Geometrics**: The visual elements used for our data, such as point, bar or line.

-   **Facets** split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).

-   **Statistics**, statiscal transformations that summarise data (e.g. mean, confidence intervals).

-   **Coordinate systems** define the plane on which data are mapped on the graphic.

-   **Themes** modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.

# Essential Grammatical Elements in ggplot2: data

Call the `ggplot()` function using the code chunk below

```{r}
ggplot(data=exam_data)
```

::: callout-note
-   A blank canvas appears.
-   `ggplot()` initializes a ggplot object.
-   The *data* argument defines the dataset to be used for plotting.
-   If the dataset is not already a data.frame, it will be converted to one by `fortify()`.
:::

# **Essential Grammatical Elements in ggplot2: [Aesthetic mappings](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html)**

The aesthetic mappings take attributes of the data and and use them to influence visual characteristics, e.g., position, colour, size, shape, or transparency. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics of a plot are specified in the [`aes()`](https://ggplot2.tidyverse.org/reference/aes.html) function call.

Code chunk below adds the aesthetic element into the plot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

::: callout-note
-   ggplot includes the x-axis and the axis’s label.
:::

# Essential Grammatical Elements in ggplot2: [geom](https://ggplot2.tidyverse.org/reference/#section-layer-geoms)

Geometric objects are the actual marks we put on a plot. Some examples:

-   *geom_point* for drawing individual points (e.g., a scatter plot)

-   *geom_line* for drawing lines (e.g., for a line charts)

-   *geom_smooth* for drawing smoothed lines (e.g., for simple trends or approximations)

-   *geom_bar* for drawing bars (e.g., for bar charts)

-   *geom_histogram* for drawing binned values (e.g. a histogram)

-   *geom_polygon* for drawing arbitrary shapes

-   *geom_map* for drawing polygons in the shape of a map! (You can access the data to use for these maps by using the map_data() function).

Note: A plot must have at least one geom; there is no upper limit. A geom can be added to a plot using the **+** operator.

## Geometric Objects: geom_bar

The code chunk below plots a bar chart by using [`geom_bar()`](https://ggplot2.tidyverse.org/reference/geom_bar.html).

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## Geometric Objects: geom_dotplot

In a dot plot, the width of a dot corresponds to the bin width (or maximum width, depending on the binning algorithm), and dots are stacked, with each dot representing one observation.

In the code chunk below, [`geom_dotplot()`](https://ggplot2.tidyverse.org/reference/geom_dotplot.html) of ggplot2 is used to plot a dot plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

::: callout-caution
The y scale is misleading and not very useful.
:::

The code chunk below does the following:

-   `scale_y_continuous()` is used to turn off the y-axis, and

-   *binwidth argument* is used to change the binwidth to 2.5.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

## **Geometric Objects: `geom_histogram()`**

The code chunk below, uses [*geom_histogram()*](https://ggplot2.tidyverse.org/reference/geom_histogram.html) create a simple histogram by using values in *MATHS* field of *exam_data*.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

::: callout-note
The default bin is **30**
:::

## **Modifying a geometric object by changing `geom()`**

The code chunk below, does the following:

-   *bins* argument is used to change the number of bins to 20,

-   *fill* argument is used to shade the histogram with light blue color, and

-   *color* argument is used to change the outline colour of the bars in black

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

## **Modifying a geometric object by changing *aes()***

The code chunk below changes the interior colour of the histogram (i.e. *fill*) by using sub-group of *aesthetic()*.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

::: callout-note
This approach can be used to colour, fill and alpha of the geometric.
:::

## **Geometric Objects: geom-density()**

[`geom-density()`](https://ggplot2.tidyverse.org/reference/geom_density.html) computes and plots [kernel density estimate](https://en.wikipedia.org/wiki/Kernel_density_estimation), which is a smoothed version of the histogram.

It is a useful alternative to the histogram for continuous data that comes from an underlying smooth distribution.

The code below plots the distribution of Maths scores in a kernel density estimate plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

The code chunk below plots two kernel density lines by using *colour* or *fill* arguments of *aes()*

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## **Geometric Objects: geom_boxplot**

[`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html) displays continuous value list. It visualises five summary statistics (the median, two hinges and two whiskers), and all “outlying” points individually.

The code chunk below plots boxplots by using [`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html).

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

[**Notches**](https://sites.google.com/site/davidsstatistics/home/notched-box-plots) are used in box plots to help visually assess whether the medians of distributions differ. If the notches do not overlap, this is evidence that the medians are different.

The code chunk below plots the distribution of Maths scores by gender in notched plot instead of boxplot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## **Geometric Objects: geom_violin**

[`geom_violin`](https://ggplot2.tidyverse.org/reference/geom_violin.html) is designed for creating violin plot. Violin plots are a way of comparing multiple data distributions. With ordinary density curves, it is difficult to compare more than just a few distributions because the lines visually interfere with each other. With a violin plot, it’s easier to compare several distributions since they’re placed side by side.

The code below plot the distribution of Maths score by gender in violin plot.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## **Geometric Objects: geom_point()**

[`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html) is especially useful for creating scatterplot.

The code chunk below plots a scatterplot showing the Maths and English grades of pupils by using `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

## ***Combining geom*** **objects**

The code chunk below plots the data points on the boxplots by using both `geom_boxplot()` and `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

# **Essential Grammatical Elements in ggplot2: stat**

The [Statistics functions](https://ggplot2.tidyverse.org/reference/#stats) statistically transform data, usually as some form of summary. For example:

-   frequency of values of a variable (bar graph)

    -   a mean

    -   a confidence limit

-   There are two ways to use these functions:

    -   add a `stat_()` function and override the default geom, or

    -   add a `geom_()` function and override the default stat.

## **Working with `stat()`**

The boxplots below are incomplete because the positions of the means were not shown.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

## **Working with stat - the *stat_summary()* method**

The code chunk below adds mean values by using [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html) function and overriding the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="red",        
               size=4)               
```

## **Working with stat - the `geom()` method**

The code chunk below adding mean values by using `geom_()` function and overriding the default stat.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun.y="mean",           
             colour ="red",          
             size=4)          
```

## **Adding a best fit curve on a scatterplot**

The interpretability of scatterplots can be improved by adding a best fit curve. n the code chunk below, [`geom_smooth()`](https://ggplot2.tidyverse.org/reference/geom_smooth.html) is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

::: callout-note
The default method used is *loess*.
:::

The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5)
```

# **Essential Grammatical Elements in ggplot2: Facets**

Facetting generates small multiples (aka trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables. ggplot2 supports two types of factes: [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) and [`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html).

## **Working with `facet_wrap()`**

[`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) wraps a 1d sequence of panels into 2d. This is generally a better use of screen space than facet_grid because most displays are roughly rectangular.

The code chunk below plots a trellis plot using `facet-wrap()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## **`facet_grid()` function**

[`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) forms a matrix of panels defined by row and column facetting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data.

The code chunk below plots a trellis plot using `facet_grid()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

# **Essential Grammatical Elements in ggplot2: Coordinates**

The *Coordinates* functions map the position of objects onto the plane of the plot. There are a number of different possible coordinate systems to use, they are:

-   [coord_cartesian()](https://ggplot2.tidyverse.org/reference/coord_cartesian.html): the default cartesian coordinate systems, where you specify x and y values (e.g. allows you to zoom in or out

-   [coord_flip()](https://ggplot2.tidyverse.org/reference/coord_flip.html): a cartesian system with the x and y flipped.

-   [coord_fixed()](https://ggplot2.tidyverse.org/reference/coord_fixed.html): a cartesian system with a "fixed" aspect ratio (e.g. 1.78 for a "widescreen" plot).

-   [coord_quickmap()](https://ggplot2.tidyverse.org/reference/coord_map.html): a coordinate system that approximates a good aspect ratio for maps.

## **Working with Coordinate**

By the default, the bar chart of ggplot2 is in vertical form.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

The code chunk below flips the horizontal bar chart into vertical bar chart by using `coord_flip()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

## **Changing the y- and x-axis range**

The scatterplot below is slightly misleading because the y-aixs and x-axis range are not equal.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

The code chunk below fixed both the y-axis and x-axis range from 0-100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

# **Essential Grammatical Elements in ggplot2: themes**

[Themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) control elements of the graph not related to the data. For example:

-   background colour

-   size of fonts

-   gridlines

-   colour of labels

Built-in themes include: - `theme_gray()` (default) - `theme_bw()` - `theme_classic()`

Each theme element can be conceived of as either a line (e.g. x-axis), a rectangle (e.g. graph background), or text (e.g. axis title).

## **Working with theme**

The code chunk below plot a horizontal bar chart using `theme_gray()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

A horizontal bar chart plotted using `theme_classic()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

A horizontal bar chart plotted using `theme_minimal()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
