---
title: "In-Class Exercise 2"
author: "Alicia Loh"
date-modified: last-modified
execute: 
  eval: true
  echo: true
  warning: false
---

# Getting Started

## Loading R packages

Note: Ensure that the [pacman](https://cran.r-project.org/web/packages/pacman/) package has already been installed.

The code chunk below uses [`p_load()`](https://www.rdocumentation.org/packages/pacman/versions/0.5.1/topics/p_load) of pacman package to load the tidyverse family of packages.

```{r}
pacman::p_load(tidyverse)
```

## Importing the Data

-   The code chunk below imports *realis2019.csv* into R environment by using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package.

-   **readr** is one of the tidyverse package.

```{r, eval=FALSE}
realis <- read_csv("data/realis2019.csv")
```

# ggplot

[ggplot2](https://ggplot2.tidyverse.org/) is an R package for **declaratively** creating **data-driven** graphics based on ***The Grammar of Graphics***. It is also part of the tidyverse family specially designed for visual exploration and communication.

```{r, eval=FALSE}
ggplot(data=realis, 
       aes(x = `Unit Price ($ psm)`)) +
  geom_histogram() +
  ggtitle("Distribution of Unit Price ($ psm)")
```
