---
title: "In-class Exercise 9 Multivariate Analysis"
author: "Alicia Loh"
date: "15 June, 2024"
date-modified: last-modified
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Installing and loading the required libraries

The following R packages will be used:

-   scatterPlotMatrix

-   parallelPlot

-   cluster

-   factoextra

-   tidyverse

Code chunk below will be used to check if these packages have been installed and also will load them into the working R environment.

```{r}
pacman::p_load(scatterPlotMatrix,parallelPlot,cluster,factoextra,tidyverse) 
```

## **Importing Data into R**

### **Importing Data**

In the code chunk below, **read_csv()** of *readr* is used to import wine_quality.csv into R and parsed it into tibble R data frame

```{r}
wine <- read_csv("data/wine_quality.csv")
```

### **View Data**

```{r}
ggplot(data= wine, 
       aes(x = type)) +
  geom_bar()
```

::: callout-note
There are 2 types of wine: Red and White
:::

## **White Wine**

Select relevant columns

```{r}
whitewine <- wine %>%
  filter(type == "white") %>%
  select(c(1:11))
```

Plot White Wine Attributes

::: panel-tabset
## Plot

```{r,echo=FALSE}
scatterPlotMatrix(whitewine,
                  corrPlotType = "Text",
                  distribType = 2,
                  width = 500,
                  height = 500,
                  rotateTitle = TRUE)
```

## Code

```{r,eval=FALSE}
scatterPlotMatrix(whitewine,
                  corrPlotType = "Text",
                  distribType = 1,
                  width = 500,
                  height = 500)
```
:::

Compute GAP Statistic

See Data Cluster (post GAP statistic)

```{r}
set.seed(1234)
kmeans4 <- kmeans(whitewine, 4, nstart=25)
print(kmeans)
```

Visualise clusters

```{r}
fviz_cluster(kmeans4, data = whitewine)
```

Populate cluster values

```{r}
whitewine <- whitewine %>%
  mutate(Cluster = kmeans4$cluster)
```

Compute cluster as factors

```{r}
whitewine$Cluster <- 
  as_factor(whitewine$Cluster)
```
